---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# HealthIneq

<!-- badges: start -->
![R-CMD-check](https://github.com/elachtara/HealthIneq/workflows/R-CMD-check/badge.svg)
<!-- badges: end -->

The goal of HealthIneq is to .shed light on the health inequities across the United States. 

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("elachtara/HealthIneq")
```
## Example

First you will load the package and call for the data. 

```{r Data read in}
library(HealthIneq)
library(dplyr)
data(HealthIneq)
```

Let's look at the life expectancy for males vs. females:

```{r Life Expectancy}
# Create average life expectancy across all counties
HealthIneq  %>% 
  summarize(male_avg = mean(avg_life_M), 
            female_avg = mean(avg_life_F))
```
Sorry fellas! Looks like females live on average over 3 years longer than males.

Let's make a plot of these life expectancies.

```{r Plot}
plot(HealthIneq$avg_life_M, HealthIneq$avg_life_F, main = "Male vs. Female Life Expectancies",
     xlab = "Male Life Expectancy (Years)", ylab = "Feale Life Expectancy (Years)")
```
Pretty neat, appears to be a linear relationship between male and female life expectancies across the counties of the United States. Check out the vingette for more examples on how to use the package.
